---
title: "Integrating Modalities: Multi-Modal Fusion with PhysioCrossModal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrating Modalities: Multi-Modal Fusion with PhysioCrossModal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## The integration problem

Movement is measured in many modalities — joint **kinematics**, **kinetics**
(forces), **EMG**, **IMU** — and the scientifically interesting question is usually
not any one of them but *what they share*. `PhysioCrossModal` provides a family of
**multi-modal fusion** methods that take several time-aligned modality blocks and
answer complementary questions:

* what is **shared across** all modalities vs **specific to** each? — `jive()`,
  `multipleFactorAnalysis()`
* which **directions co-vary** between two modalities? — `cca()`, `plsBlocks()`
* is there a **common latent drive**? — `coupledNMF()`
* how related are the modalities **as a whole**? — `rvCoefficient()`,
  `distanceCorrelation()`, `representationalSimilarity()`
* one composite **index** / design matrix / latent model — `multimodalGaitIndex()`,
  `fuseBlocks()`, `crossBlockFactor()`

Every method takes a list of matrices that **share the observation dimension**
(rows = the same subjects, or gait-cycle points, or time samples), with each
modality contributing its own columns (features).

```{r data}
library(PhysioCrossModal)

# two aligned modality blocks: rows = the same 200 observations (e.g. subjects),
# columns = each modality's features
kinematics <- matrix(rnorm(200 * 12), 200, 12)   # e.g. joint-angle features
kinetics   <- matrix(rnorm(200 *  8), 200,  8)   # e.g. ground-reaction-force features
```

## Shared vs modality-specific: JIVE

**JIVE** (Joint and Individual Variation Explained) splits each block into a
low-rank **joint** structure common to *all* modalities, low-rank **individual**
structure specific to each, and residual. The joint scores are the *integrated
representation*; the variance table tells you how much of each modality is shared.

```{r jive}
fit <- jive(list(kinematics = kinematics, kinetics = kinetics),
            rank_joint = 2, rank_individual = 2)

fit$variance          # joint / individual / residual fraction per modality
fit$joint_scores      # the shared component scores (integrated representation)

# relate the shared component to an outcome (e.g. walking speed):
# cor(fit$joint_scores[, 1], speed)
```

**Multiple Factor Analysis** is a block-balanced global PCA — each block is scaled
so a large or high-variance modality cannot dominate the shared factors.

```{r mfa}
mfa <- multipleFactorAnalysis(list(kinematics, kinetics), ncomp = 3)
mfa$explained         # variance per global factor
mfa$contributions     # each block's contribution to each factor
```

## Covarying directions: CCA and PLS

**Canonical correlation** finds the linear combinations of two modalities whose
scores are maximally *correlated* (auto-regularised when features exceed
observations); **two-block PLS** maximises *covariance* instead, and is robust in
high dimensions.

```{r cca}
cc <- cca(kinematics, kinetics)
cc$cor                # canonical correlations (decreasing)
cc$xscores; cc$yscores

pl <- plsBlocks(kinematics, kinetics, ncomp = 2)
pl$correlation        # score correlation per component
```

## A common latent drive: coupled NMF

When two **non-negative** modalities share a latent drive (e.g. EMG envelopes and a
kinematic-speed envelope driven by the same temporal command), `coupledNMF()`
estimates one shared factor `W` with modality-specific loadings.

```{r coupled}
emg    <- abs(matrix(rnorm(200 * 6), 200, 6))   # non-negative
imu    <- abs(matrix(rnorm(200 * 4), 200, 4))
cn <- coupledNMF(emg, imu, rank = 2)
cn$shared             # W: the common drive
cn$vaf_x; cn$vaf_y    # fit per modality
```

## Whole-block association

`rvCoefficient()` is a multivariate correlation between two blocks;
`distanceCorrelation()` also captures **non-linear** dependence (zero iff
independent); `representationalSimilarity()` compares each modality's *geometry of
observations* (RSA). All offer permutation p-values.

```{r assoc}
rvCoefficient(kinematics, kinetics, permutations = 1000)$rv
distanceCorrelation(kinematics, kinetics, permutations = 1000)$dcor
representationalSimilarity(list(kin = kinematics, kinet = kinetics))$similarity
```

A low RV with a high canonical correlation means the coupling is real but
*low-dimensional* — read several lenses together.

## One index, one matrix, one latent model

A **composite multi-modal gait index** (Gait-Deviation-Index style, but spanning
modalities) scores how far a subject's multi-modal feature vector lies from a
normative reference, with a per-modality breakdown:

```{r index}
reference <- matrix(rnorm(300 * 20), 300, 20)     # controls
subjects  <- matrix(rnorm(10 * 20), 10, 20)
blocks    <- list(kinematics = 1:8, kinetics = 9:14, emg = 15:20)
idx <- multimodalGaitIndex(subjects, reference, blocks = blocks)
idx$index          # ~100 = typical; lower = more deviant
idx$by_modality    # which modality drives each subject's deviation
```

`fuseBlocks()` produces a single block-balanced design matrix for downstream
multi-view machine learning, and `crossBlockFactor()` fits a common-factor model
that flags which latent factors are **shared across modalities** vs
modality-specific.

```{r fuse}
fused <- fuseBlocks(list(kinematics, kinetics), method = "mfa")$matrix
cf <- crossBlockFactor(list(kinematics, kinetics), nfactors = 2)
cf$shared          # TRUE for factors that load on >= 2 modalities
```

## Choosing a method

| question | method |
|---|---|
| what is shared vs modality-specific? | `jive()` |
| block-balanced shared factors | `multipleFactorAnalysis()` |
| which directions co-vary? (correlation / covariance) | `cca()` / `plsBlocks()` |
| a common non-negative drive | `coupledNMF()` |
| whole-block association (linear / non-linear / geometry) | `rvCoefficient()` / `distanceCorrelation()` / `representationalSimilarity()` |
| one composite deviation score | `multimodalGaitIndex()` |
| feature-level fusion for ML | `fuseBlocks()` |
| shared vs specific latent factors | `crossBlockFactor()` |

## Correctness

These methods are **cross-validated against independent, established
implementations** — `cca()` vs base `cancor`, `distanceCorrelation()` vs
`energy::dcor`, `rvCoefficient()` vs `FactoMineR::coeffRV`,
`multipleFactorAnalysis()` vs `FactoMineR::MFA`, `representationalSimilarity()` vs
`vegan::mantel`, `crossBlockFactor()` vs `stats::factanal`, and `jive()` against its
exact algebraic identity — agreeing to machine precision (see
`planning/crossmodal-validation/` in the development monorepo), in addition to the
packaged unit tests.

## See also

* `vignette("coupling-analysis", package = "PhysioCrossModal")` — pairwise
  cross-modal coupling (coherence, phase, PAC, Granger).
* The `MultiPhysioExperiment` container for storing and aligning modalities.
